Adaptive Approximation for Multivariate Linear Problems with Inputs Lying in a Cone

### 2019-09-16 DingHickKritzMakRICAMChapVerD.tex
Updates from Overleaf

In Sections \ref{DHKM:pilot_sec} and \ref{DHKM:tracking_sec}, the weights $\bslambda = (\lambda_{\bsk_1}, \lambda_{\bsk_2}, \ldots )$, which appear in the definition of the cone of inputs, $\calC_{\bslambda,n_1,A}$ and $\calC_{\bslambda,\bsn,a,b}$, are taken as given and fixed.  One may assume the form suggested in \eqref{DHKM:prodwts}, which defines $\bslambda$ in terms of coordinate weights, $\bsw$, and smoothness weights, $\bss$.  However, practically speaking it may be difficult to know a priori suitable values for these weights.
This section explores a situation where the initial data collected for the input function data can be used to learn $\bsw$ and $\bss$, inferring which input variables in $f$ may be more important and the smoothness of the function.

@@

\subsection{Product, Order and Smoothness Dependent (POSD) Weights}
The $u_{\bsk}$ and the $\lambda_{\bsk}$ in this section are defined as 
\begin{equation}
u_\bsk = \prod_{\ell = 1}^d \tu_{k_\ell}, \quad 
\lambda_{\bsk} = \Gamma_{\|\bsk\|_0} \prod_{\substack{\ell=1\\ k_\ell>0}}^d w_\ell s_{k_\ell}, \quad \Gamma_0 = s_1 = 1, \quad \bsk \in \mathbb{N}_0^d,
\label{DHKM:posdeq}
\end{equation}
where $\bsw = (w_\ell)_{l=1}^d$ is the vector of coordinate weights, $\bss = (s_k)_{k=1}^\infty$ is the vector of smoothness weights,  $\boldsymbol{\Gamma} = (\Gamma_m)_{m=1}^d$ is the vector of \emph{order weights}, and $\norm[0]{\bsk}$ denotes the number of nonzero elements of $\bsk$. The intuition behind these weights is as follows: 
\begin{itemize}
    \item Coordinate weights quantify the importance for the $d$ input variables in $f$.
    \item Smoothness weights quantify the importance of the $\tu_k$.  E.g., if the $\tu_k$ are polynomials of degree $k$ as in Section \ref{DHKM:secexamp}, then the faster the $s_k$ decay, the smoother $f$ is.  
    \item Order weights quantify the importance of effects with different orders; $\bsk$ having one nonzero element corresponds to a first-order or main effect, $\bsk$ having two nonzero elements corresponds to a second-order (interaction) effect.   (e.g., first-order, second-order).
\end{itemize} \ \vspace{-4ex}

This parametrization is motivated by several guiding principles from the experimental design literature \citep{WuHam2009}, which are briefly described below.  In statistical parlance, the terms $\hf(\bsk) u_{\bsk}$ are effects.
\begin{itemize}
\item \emph{Effect sparsity} assumes that only a small number of \DHKMchange{coordinates in} $f$ are important. In \eqref{DHKM:posdeq}, this sparsity means that only a small number of product weights $\bsw$ are large.  This principle arises in the sufficient condition for strong tractability in Corollary \ref{DHKM:sptexample_cor}.
\item \emph{Effect heredity} assumes that lower-order effects are more important than higher-order effects. E.g., $\lambda_{(1,0, 0,\ldots, 0)}$ should be larger than $\lambda_{(1, 1, 0, \ldots, 0)}$. In \eqref{DHKM:posdeq}, this heredity can be enforced by assuming that the order weights $\Gamma_{m}$ decrease with $m$. 
\item \emph{Effect hierarchy} assumes that an effect is active \emph{only} when all its component effects are active. For example, $\lambda_{(1, 1, 0, \ldots, 0)} > 0$ only when $\lambda_{(1, 0, 0, \ldots, 0)}$ and $\lambda_{(0, 1, 0, \ldots, 0)}$ are both nonzero. This hierarchy is implicitly enforced by the product structure of the weights in \eqref{DHKM:posdeq}.
\item \emph{Effect smoothness} assumes that lower-degree effects are more important than higher-degree effects. For example, when the $(\tu_k)_{k \in \mathbb{N}_0}$ are polynomials, this means that linear effects are more important than quadratic effects, which are in turn more significant than cubic effects, and so on. Effect smoothness can be imposed by assuming $\bss$ to be a decreasing sequence. 
\end{itemize} 
\vspace{-3ex}
\DHKMchange{These principles will guide the inference of weights $\bslambda$ (and its corresponding cone of inputs $\calC_{\bslambda,A}$) from initial data.}

@@

%Let $\calC_{\bslambda}$ denote the cone of inputs defined in \eqref{DHKM:pilot_cone} by POSD weights $\bslambda = \bigl( \lambda_{\bsk} \bigr)_{\bsk \in \N_0^d}$.
As mentioned above, our goal here is to infer $\bslambda$ from input function data.  We start with an initial set of wavenumbers:
\begin{equation} \label{DHKM:barKdef}
    \bcalK= \{ (0, \ldots, 0, k, 0, \ldots, 0): k = 0, \ldots, k_{\max}\},
\end{equation}
\DHKMchange{where the only non-zero entry $k$ may be at any position in the vector}. The approximation to $f$ based on sampling the series coefficients for these wavenumbers is
\begin{equation*}
    \fapp = \sum_{\bsk \in \bar{\calK}} \hf(\bsk) u_{\bsk}.
\end{equation*}
\DHKMchange{To choose the cone $\calC_{\bslambda,A}$ which best fits $f$, we select the POSD weights $\bslambda$ to make the norm of $f_{\text{app}}$ small. These weights can be written as:}
\begin{multline}
\bar{\bslambda} = \bslambda(\bar{\bsw}, \bar{\bss}, \bsGamma), \\
\text{where } (\bar{\bsw}, \bar{\bss} ) = \min \left\{ \argmin_{(\bsw,\bss) \in \calW \times \calS} \left\|\left( \frac{\hf(\bsk)}{\lambda_{\bsk}(\bsw,\bss,\bsGamma)} \right)_{\bsk \in \bar{\calK}}\right\|_{\rho} \right \}.
\label{DHKM:eq:inf}
\end{multline}
Here, $\calW$ is a candidate set for coordinate weights, e.g., $\calW = [0,w^*]^d$, and $\calS$ is a candidate set for the smoothness weights, e.g., $\calS = \{(1/k^r)_{k=1}^{\infty} \colon r > 0\}$. The inner minimization finds the $(\bsw,\bss)$ that minimizes the approximate norm of the input function.  This minimizer may be non-unique, so the outer minimization chooses the smallest such $(\bsw,\bss)$. Making the coordinate and smoothness weights as small as possible helps enforce the principles of effect sparsity.  The optimum, $(\bar{\bsw}, \bar{\bss})$, then defines the \emph{data-inferred} POSD $\bslambda$, denoted $\bar{\bslambda}$.

@@

Algorithm \ref{DHKM:InfPilotConeAlg} combines the construction of data-inferred POSD weights, $\bar{\bslambda}$, with Algorithm \ref{DHKM:PilotConeAlg} of Section \ref{DHKM:pilot_sec}. This algorithm succeeds for input functions in the cone 
\begin{equation} \label{DHKM:pilot_cone2}
     \DHKMchange{\bar{\calC}_{A}} := \left \{ f \in \calF : f \in \DHKMchange{\calC_{\bar{\bslambda},A}} \text{ for } \bar{\bslambda} \text{ defined in \eqref{DHKM:eq:inf}} \right\}.
\end{equation}
The reason that \DHKMchange{$\bar{\calC}_A$} is a cone is that the data-inferred $\bar{\bslambda}$ for the input function $f$ is exactly the same as for the input function $cf$, where $c$ is any constant.

@@

\Ensure Error criterion \eqref{DHKM:err_crit} for the cone defined in \eqref{DHKM:pilot_cone2}
\State  Define the initial set of wavenumbers $\bcalK$ defined in \eqref{DHKM:barKdef}
\State Evaluate initial sample $\bigl\{\hf(\bsk)\bigr\}_{\bsk \in \bcalK}$
\State Compute data-driven POSD weight $\bar{\bslambda}$\DHKMchange{, by iteratively optimizing over $\bsw \in \calW$ and $\bss \in \calS$ in \eqref{DHKM:eq:inf} until convergence}
\State Using these weights, $\bar{\bslambda}$, perform Algorithm \ref{DHKM:PilotConeAlg} to obtain $\ALG(f,\varepsilon)$
\RETURN $\ALG(f,\varepsilon)$
\end{algorithmic}
\end{algorithm}

>>>>

In Sections \ref{DHKM:pilot_sec} and \ref{DHKM:tracking_sec}, the weights \DHKMchange{$\bslambda = (\lambda_{\bsk_1}, \lambda_{\bsk_2}, \ldots )$}, which appear in the definition of the cone of inputs, \DHKMchange{$\calC_{\bslambda,n_1,A}$ and $\calC_{\bslambda,\bsn,a,b}$, are taken as given and fixed.  One may assume the form suggested in \eqref{DHKM:prodwts}, which defines $\bslambda$ in terms of coordinate weights, $\bsw$, and smoothness weights, $\bss$.  However, practically speaking it may be difficult to know a priori suitable values for these weights.
This section explores a situation where the initial data collected for the input function data can be used to learn $\bsw$ and $\bss$, inferring which input variables in $f$ may be more important and the smoothness of the function.  We focus on inferring the $\bslambda$ defining the cone $\calC_{\bslambda,n_1,A}$ in Section \ref{DHKM:pilot_sec}.}

@@

\subsection{Product, Order and Smoothness Dependent (POSD) Weights}
The $u_{\bsk}$ and the $\lambda_{\bsk}$ in this section are defined as 
\begin{equation}
u_\bsk = \prod_{\ell = 1}^d \tu_{k_\ell}, \quad 
\lambda_{\bsk} = \Gamma_{\|\bsk\|_0} \prod_{\substack{\ell=1\\ k_\ell>0}}^d w_\ell s_{k_\ell}, \quad \Gamma_0 = s_1 = 1, \quad \bsk \in \mathbb{N}_0^d,
\label{DHKM:posdeq}
\end{equation}
where $\bsw = (w_\ell)_{l=1}^d$ is the vector of coordinate weights, $\bss = (s_k)_{k=1}^\infty$ is the vector of smoothness weights,  $\boldsymbol{\Gamma} = (\Gamma_m)_{m=1}^d$ is the vector of \emph{order weights}, and $\norm[0]{\bsk}$ denotes the number of nonzero elements of $\bsk$. \DHKMchange{This is a generalization of the form for $\bslambda$ given by \eqref{DHKM:prodwts}.} The intuition behind these weights is as follows: 
\begin{itemize}
    \item Coordinate weights quantify the importance for the $d$ input variables in $f$.
    \item Smoothness weights quantify the importance of the $\tu_k$.  E.g., if the $\tu_k$ are polynomials of degree $k$ as in Section \ref{DHKM:secexamp}, then the faster the $s_k$ decay, the smoother $f$ is.  
    \item Order weights quantify the importance of effects with different orders; $\bsk$ having one nonzero element corresponds to a first-order or main effect, $\bsk$ having two nonzero elements corresponds to a second-order (interaction) effect.   (e.g., first-order, second-order).
\end{itemize} \ \vspace{-4ex}

This parametrization is motivated by several guiding principles from the experimental design literature \citep{WuHam2009}, which are briefly described below.  In statistical parlance, the terms $\hf(\bsk) u_{\bsk}$ are effects.
\begin{itemize}
\item \emph{Effect sparsity} assumes that only a small number of \DHKMchange{coordinates in} $f$ are important. In \eqref{DHKM:posdeq}, this sparsity means that only a small number of product weights $\bsw$ are large.  This principle arises in the sufficient condition for strong tractability in Corollary \ref{DHKM:sptexample_cor}.
\item \emph{Effect heredity} assumes that lower-order effects are more important than higher-order effects. E.g., $\lambda_{(1,0, 0,\ldots, 0)}$ should be larger than $\lambda_{(1, 1, 0, \ldots, 0)}$. In \eqref{DHKM:posdeq}, this heredity can be enforced by assuming that the order weights $\Gamma_{m}$ decrease with $m$. 
\item \emph{Effect hierarchy} assumes that an effect is active \emph{only} when all its component effects are active. For example, $\lambda_{(1, 1, 0, \ldots, 0)} > 0$ only when $\lambda_{(1, 0, 0, \ldots, 0)}$ and $\lambda_{(0, 1, 0, \ldots, 0)}$ are both nonzero. This hierarchy is implicitly enforced by the product structure of the weights in \eqref{DHKM:posdeq}.
\item \emph{Effect smoothness} assumes that lower-degree effects are more important than higher-degree effects. For example, when the $(\tu_k)_{k \in \mathbb{N}_0}$ are polynomials, this means that linear effects are more important than quadratic effects, which are in turn more significant than cubic effects, and so on. Effect smoothness can be imposed by assuming $\bss$ to be a decreasing sequence. 
\end{itemize} 
\vspace{-3ex}
\DHKMchange{These principles will guide the inference of weights $\bslambda$ (and its corresponding cone of inputs $\calC_{\bslambda,n_1,A}$ from initial data.}

@@

%Let $\calC_{\bslambda}$ denote the cone of inputs defined in \eqref{DHKM:pilot_cone} by POSD weights $\bslambda = \bigl( \lambda_{\bsk} \bigr)_{\bsk \in \N_0^d}$.
As mentioned above, our goal here is to infer $\bslambda$ from input function data.  We start with an initial set of \DHKMchange{$d k_{\max} + 1$} wavenumbers:
\begin{equation} \label{DHKM:barKdef}
    \bcalK= \{ (0, \ldots, 0, k, 0, \ldots, 0): k = 0, \ldots, k_{\max}\},
\end{equation}
\DHKMchange{where the possible non-zero entry $k$ may be at any one position in the vector}. The approximation to $f$ based on sampling the series coefficients for these wavenumbers is
\begin{equation*}
    \fapp = \sum_{\bsk \in \bar{\calK}} \hf(\bsk) u_{\bsk}.
\end{equation*}
\DHKMchange{To choose the cone $\calC_{\bslambda,n_1,A}$ which best fits $f$, we select the POSD weights $\bslambda$ to make the norm of $f_{\text{app}}$ small. These weights can be written as:}
\begin{multline}
\bar{\bslambda} = \bslambda(\bar{\bsw}, \bar{\bss}, \bsGamma), \\
\text{where } (\bar{\bsw}, \bar{\bss} ) = \min \left\{ \argmin_{(\bsw,\bss) \in \calW \times \calS} \left\|\left( \frac{\hf(\bsk)}{\lambda_{\bsk}(\bsw,\bss,\bsGamma)} \right)_{\bsk \in \bar{\calK}}\right\|_{\rho} \right \}.
\label{DHKM:eq:inf}
\end{multline}
Here, $\calW$ is a candidate set for coordinate weights, e.g., $\calW = [0,w^*]^d$, and $\calS$ is a candidate set for the smoothness weights, e.g., $\calS = \{(1/k^r)_{k=1}^{\infty} \colon r > 0\}$. The inner minimization finds the $(\bsw,\bss)$ that minimizes the approximate norm of the input function.  This minimizer may be non-unique, so the outer minimization chooses the smallest such $(\bsw,\bss)$. Making the coordinate and smoothness weights as small as possible helps enforce the principles of effect sparsity.  The optimum, $(\bar{\bsw}, \bar{\bss})$, then defines the \emph{data-inferred} POSD $\bslambda$, denoted $\bar{\bslambda}$.

@@

Algorithm \ref{DHKM:InfPilotConeAlg} combines the construction of data-inferred POSD weights, $\bar{\bslambda}$, with Algorithm \ref{DHKM:PilotConeAlg} of Section \ref{DHKM:pilot_sec}. This algorithm succeeds for input functions in the cone 
\begin{equation} \label{DHKM:pilot_cone2}
     \DHKMchange{\bar{\calC}_{n_1,A}} := \left \{ f \in \calF : f \in \DHKMchange{\calC_{\bar{\bslambda},n_1,A}} \text{ for } \bar{\bslambda} \text{ defined in \eqref{DHKM:eq:inf}} \right\}.
\end{equation}
The reason that \DHKMchange{$\bar{\calC}_{n_1,A}$} is a cone is that the data-inferred $\bar{\bslambda}$ for the input function $f$ is exactly the same as for the input function $cf$, where $c$ is any constant.

@@

\Ensure Error criterion \eqref{DHKM:err_crit} for the cone defined in \eqref{DHKM:pilot_cone2}
\State  Define the initial set of wavenumbers $\bcalK$ defined in \eqref{DHKM:barKdef}
\State Evaluate initial sample $\bigl\{\hf(\bsk)\bigr\}_{\bsk \in \bcalK}$
\State Compute data-driven POSD weight $\bar{\bslambda}$\DHKMchange{, by iteratively optimizing over $(\bsw, \bss) \in \calW \times \calS$ in \eqref{DHKM:eq:inf} until convergence}
\State Using these weights, $\bar{\bslambda}$, perform Algorithm \ref{DHKM:PilotConeAlg} to obtain $\ALG(f,\varepsilon)$
\RETURN $\ALG(f,\varepsilon)$
\end{algorithmic}
\end{algorithm}
